---
title: "Highschool and Beyond Survey Analysis"
author: "Joshua Xavier"
date: "2026-09-14"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    fig_width: 4
    fig_caption: yes
    number_sections: yes
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    theme: lumen
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
editor_options: 
  chunk_output_type: inline
---

```{=html}

<style type="text/css">

/* Cascading Style Sheets (CSS) is a stylesheet language used to describe the presentation of a document written in HTML or XML. it is a simple mechanism for adding style (e.g., fonts, colors, spacing) to Web documents. */

h1.title {  /* Title - font specifications of the report title */
  font-size: 24px;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}
h4.author { /* Header 4 - font specifications for authors  */
  font-size: 20px;
  font-family: system-ui;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - font specifications for the date  */
  font-size: 18px;
  font-family: system-ui;
  color: DarkBlue;
  text-align: center;
}
h1 { /* Header 1 - font specifications for level 1 section title  */
    font-size: 22px;
    font-family: system-ui;
    color: navy;
    text-align: left;
}
h2 { /* Header 2 - font specifications for level 2 section title */
    font-size: 20px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - font specifications of level 3 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - font specifications of level 4 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

</style>
```
```{r setup, include=FALSE}
# Detect, install and load packages if needed.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("MASS")) {
   install.packages("MASS")
   library(MASS)
}
if (!require("car")) {
   install.packages("car")
   library(car)
}
if (!require("scales")) {
install.packages("scales")                                        
library("scales") 
}
#
# specifications of outputs of code in code chunks
knitr::opts_chunk$set(echo = TRUE,      # include code chunk in the output file
                      warnings = FALSE, # sometimes, you code may produce warning messages,
                                        # you can choose to include the warning messages in
                                        # the output file. 
                      messages = FALSE, #
                      results = TRUE,   # you can also decide whether to include the output
                                        # in the output file.
                      comment = NA      # remove leading ## in the model output
                      )   
library(tidyverse)
library(knitr)
library(tidyr)
library(psych)
library(VIM)
library(dplyr)
library(pander)
library(broom)

```

\

# Data set and Practical Questions
In this section we will go over the data set, practical questions asked about the data, and preform exploratory data analysis to create the best model of the data. 


## Data Description

The data set used in this project is the "hsb2" data set, an acronym for "high school and beyond". The data set contains 200 randomly selected observations from the "High School and Beyond" survey, which was conducted by the National Center for Education Statistics and polled seniors in high school. The dataset uses 11 total variables, 1 identifying variable (id), 5 numerical variables, including the response variable write, and 5 categorical variables.

The dataset contains enough information to investigate these questions because it includes socioeconomic status, academic program, demographic characteristics, and multiple standardized test scores for 200 students, with 11 variables. This allows us to compare academic preformance across different groups and use statistical methods to determine whether a meaningful relationships exist.

* id: student id number
* write (Y): writing test score
* read (X1): reading test score
* math (X2): math test score
* science (X3): science test score
* socst (X4): social studies test score
* gender: student gender (2 levels)
* race: student race/ethnicity (4 levels)
* ses: socioeconomic status - low/middle/high (3 levels)
* schtyp: school type - public/private (2 levels)
* prog: program type - general/academic/vocational (3 levels)

```{r import-data}
hsb2 <- read.csv("https://raw.githubusercontent.com/Joshuaxav/STA321F26/main/hsb2.csv")
hsb2$gender <- factor(hsb2$gender)
hsb2$race   <- factor(hsb2$race)
hsb2$ses    <- factor(hsb2$ses, levels = c("low", "middle", "high"))
hsb2$schtyp <- factor(hsb2$schtyp)
hsb2$prog   <- factor(hsb2$prog)
```

## Practical Question

The primary question is to identify the association between a student's
writing score and other test scores available in the data set, and to
build a simple linear regression model that captures this relationship.



## Exploratory Data Analysis

We first explore the pairwise association between the numerical
variables in the data set.

```{r pairwise-scatterplot, fig.align='center', fig.width=7, fig.height=7, fig.cap="Figure 1: Pairwise scatter plot of numerical variables."}
pairs.panels(hsb2[, c("read", "write", "math", "science", "socst")],
             main = "Figure 1: Pairwise Scatterplots of Numerical Variables",
             pch = 21,
             bg = "lightblue")
```