---
title: "Highschool and Beyond Survey Analysis"
author: "Joshua Xavier"
date: "2026-09-14"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    fig_width: 4
    fig_caption: yes
    number_sections: yes
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    theme: lumen
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
editor_options: 
  chunk_output_type: inline
---

```{=html}

<style type="text/css">

/* Cascading Style Sheets (CSS) is a stylesheet language used to describe the presentation of a document written in HTML or XML. it is a simple mechanism for adding style (e.g., fonts, colors, spacing) to Web documents. */

h1.title {  /* Title - font specifications of the report title */
  font-size: 24px;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}
h4.author { /* Header 4 - font specifications for authors  */
  font-size: 20px;
  font-family: system-ui;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - font specifications for the date  */
  font-size: 18px;
  font-family: system-ui;
  color: DarkBlue;
  text-align: center;
}
h1 { /* Header 1 - font specifications for level 1 section title  */
    font-size: 22px;
    font-family: system-ui;
    color: navy;
    text-align: left;
}
h2 { /* Header 2 - font specifications for level 2 section title */
    font-size: 20px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - font specifications of level 3 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - font specifications of level 4 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

</style>
```
```{r setup, include=FALSE}
# Detect, install and load packages if needed.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("MASS")) {
   install.packages("MASS")
   library(MASS)
}
if (!require("car")) {
   install.packages("car")
   library(car)
}
if (!require("scales")) {
install.packages("scales")                                        
library("scales") 
}
#
# specifications of outputs of code in code chunks
knitr::opts_chunk$set(echo = TRUE,      # include code chunk in the output file
                      warnings = FALSE, # sometimes, you code may produce warning messages,
                                        # you can choose to include the warning messages in
                                        # the output file. 
                      messages = FALSE, #
                      results = TRUE,   # you can also decide whether to include the output
                                        # in the output file.
                      comment = NA      # remove leading ## in the model output
                      )   
library(tidyverse)
library(knitr)
library(tidyr)
library(psych)
library(VIM)
library(dplyr)
library(pander)
library(broom)

```

\

# Data set and Practical Questions
In this section we will go over the data set, practical questions asked about the data, and preform exploratory data analysis to create the best model of the data. 


## Data Description

The data set used in this project is the "hsb2" data set, an acronym for "high school and beyond". The data set contains 200 randomly selected observations from the "High School and Beyond" survey, which was conducted by the National Center for Education Statistics and polled seniors in high school. The dataset uses 11 total variables, 1 identifying variable (id), 5 numerical variables, including the response variable write, and 5 categorical variables.

The dataset contains enough information to investigate these questions because it includes socioeconomic status, academic program, demographic characteristics, and multiple standardized test scores for 200 students, with 11 variables. This allows us to compare academic preformance across different groups and use statistical methods to determine whether a meaningful relationships exist.

* id: student id number
* write (Y): writing test score
* read (X1): reading test score
* math (X2): math test score
* science (X3): science test score
* socst (X4): social studies test score
* gender: student gender (2 levels)
* race: student race/ethnicity (4 levels)
* ses: socioeconomic status - low/middle/high (3 levels)
* schtyp: school type - public/private (2 levels)
* prog: program type - general/academic/vocational (3 levels)

```{r import-data}
hsb2 <- read.csv("https://raw.githubusercontent.com/Joshuaxav/STA321F26/main/hsb2.csv")
hsb2$gender <- factor(hsb2$gender)
hsb2$race   <- factor(hsb2$race)
hsb2$ses    <- factor(hsb2$ses, levels = c("low", "middle", "high"))
hsb2$schtyp <- factor(hsb2$schtyp)
hsb2$prog   <- factor(hsb2$prog)
```

## Practical Question

The primary question is to identify the association between a student's
writing score and other test scores available in the data set, and to
build a simple linear regression model that captures this relationship.



## Exploratory Data Analysis

We first explore the pairwise association between the numerical
variables in the data set.

```{r pairwise-scatterplot, fig.align='center', fig.width=7, fig.height=7, fig.cap="Figure 1: Pairwise scatter plot of numerical variables."}
pairs.panels(hsb2[, c("read", "write", "math", "science", "socst")],
             main = "Figure 1: Pairwise Scatterplots of Numerical Variables",
             pch = 21,
             bg = "lightblue")
```
The five variables all show a positive relationship with each other. In general, students who scored higher in one subject also scored higher in the others. Reading and math had the strongest relationship at 0.66, while science and social studies had the weakest at 0.47. It is also worth mentioning reading had a particularly strong relationship with the other subjects. 

The points follow a fairly straight upward pattern, so a linear model appears to be appropriate for predicting writing scores from reading scores.

Because `read`, `math`, `science`, and `socst` are all correlated with
one another (0.47–0.66), we will need to check for multicollinearity
once they're all in the same model together.

If a transformation is needed, we would transform the response variable. This is because
the Box Cox transformation helps fix stabilize variance, normalize residuals, and improves linearity, which are tied to the response variable in a linear regression model. 

# Fitting MLR to Data
We use parametric and bootstrap regression models to assess the
association between writing score and reading score

## Raw Model and Diagnostics
We start with a model that includes every available predictor.

```{r raw-model, fig.align='center', fig.cap="Figure 2"}
full.model <- lm(write ~ read + math + science + socst +
                    gender + race + ses + schtyp + prog, data = hsb2)
kable(summary(full.model)$coef, digits = 4,
      caption = "Statistics of Regression Coefficients - Full Model")
```

## Multicollinearity Check (VIF)

Since the four numerical predictors are correlated with each other, we
check the variance inflation factor before trusting any individual
coefficient's p-value.

```{r full-vif, fig.align='center', fig.width=7, fig.height=5,fig.cap="Figure 2: VIF Values" }
vif(full.model)
barplot(vif(full.model)[, 1], main = "Figure 2: VIF Values", horiz = FALSE, col = "steelblue")
        
```

All VIFs are below 2.5, well under the threshold of 4-10,
so multicollinearity is not a concern here despite the correlation seen
in the pairwise plot.

Next, we conduct residual diagnostic analysis to check the validity of
the model before making an inference about the model.

```{r raw-diagnostics, fig.align='center', fig.width=5, fig.height=5, fig.cap="Figure 3: Residual plots of the full model."}
par(mfrow = c(2, 2))
plot(full.model, main = "Figure 3: Residual plots of full model")
tidy_model <- tidy(full.model)
kable(tidy_model, digits = 4,
      caption = paste0("Coefficient estimates (R² = ",
                        round(summary(full.model)$r.squared, 3), ")"))
```

The residual plots for the full model look reasonably good already, with
no strong curve pattern and no obvious funnel shape. We revisit these
diagnostics more formally after variable selection, since removing
predictors can change the residual structure.

Looking at the full-model coefficient table above, `race`, `ses`,
`schtyp`, and `prog` are all statistically insignificant with p-values
between 0.19 and 0.66, while `read`, `math`, `science`, `socst`, and
`gender` are all significant or borderline significant. Since the VIF
check ruled out multicollinearity as an explanation, we drop the four
insignificant categorical predictors and refit.

```{r reduced-model, fig.align='center', fig.cap="Figure 4: Reduced Mofel Residual Plots}
reduced.model <- lm(write ~ read + math + science + socst + gender, data = hsb2)
kable(summary(reduced.model)$coef, digits = 4,
      caption = "Statistics of Regression Coefficients - Reduced Model")
par(mfrow = c(2, 2))
plot(reduced.model, main = "Figure 4: Reduced Model Residual Plots")
shapiro.test(residuals(reduced.model))

```
With `race`, `ses`, `schtyp`, and `prog` removed, the reduced model
keeps `read`, `math`, `science`, `socst`, and `gender`. R² drops only
slightly, from 0.622 to 0.602, for a much simpler model with far fewer
coefficients.

The Q-Q plot tracks the reference line closely except for two tail
points, and the Shapiro-Wilk test returns p ≈ 0.975, giving no evidence
against normality of the residuals. The Residuals vs Fitted and
Scale-Location plots, however, both show a visible bend in the red
smoother line rather than a perfectly flat trend. Since a plain
normality test doesn't tell us whether that bend is meaningful, we check
it directly with a Box-Cox transformation below.


## Box-Cox Transformation Check

```{r boxcox-mlr-plot, fig.align='center', fig.width=5, fig.height=5, fig.cap="Figure 5: Box-Cox transformation of the reduced model."}
boxcox(reduced.model, lambda = seq(-1, 4, 0.1))
```

The Box-Cox plot suggests an optimal $\lambda \approx 2.1$, with an
approximate 95% confidence interval of about (1.5, 2.7) - this does not
include 1, so a transformation is technically indicated even though the
Shapiro-Wilk test on the raw residuals looked fine. We check this
directly by applying $\lambda = 2.1$.

```{r boxcox-mlr-apply, fig.align='center', fig.width=7, fig.height=7, fig.cap="Figure6: Residual plots of the Box-Cox transformed model."}
write <- hsb2$write
lambda   <- 2.1
write.bc <- (write^lambda - 1) / lambda
boxcox.model <- lm(write.bc ~ read + math + science + socst + gender, data = hsb2)

par(mfrow = c(2, 2))
plot(boxcox.model, main = "Figure 6: Residual Plots of Box-Cox transformed model")
```

```{r boxcox-mlr-compare}
compare_table <- tibble(
  Model = c("write ~ ... (raw)", "Box-Cox transformed (lambda = 2.1)"),
  R_squared = c(summary(reduced.model)$r.squared, summary(boxcox.model)$r.squared),
  Shapiro_p = c(shapiro.test(residuals(reduced.model))$p.value,
                shapiro.test(residuals(boxcox.model))$p.value)
)
kable(compare_table, digits = 4, caption = "Raw vs. Box-Cox transformed model comparison")
```

The transformed model shows almost no improvement. R² moves from 0.602
to 0.605, and the Shapiro-Wilk p-value actually decreases slightly, from
0.975 to 0.937, though both remain far from significant. More
importantly, the bend visible in the Residuals vs Fitted and
Scale-Location plots is essentially unchanged after transformation.
This tells us the bend is not an issue that a power
transformation can correct, but more likely reflects some
nonlinearity in how one of the predictors relates to writing score.
Given the negligible gain, **we retain the untransformed model**
as our final model.

## Goodness-of-Fit Comparison

Next, we extract several other goodness-of-fit from 2 candidate models and summarize them in the following table.

Since our final decision is between the full and reduced models, both
on the original scale, SSE, AIC, SBC, and PRESS are all valid to
compare directly.

```{r select-function}
select <- function(m) {
  e   <- m$resid
  n0  <- length(e)
  SSE <- (m$df) * (summary(m)$sigma)^2
  R.sq  <- summary(m)$r.squared
  R.adj <- summary(m)$adj.r
  MSE   <- (summary(m)$sigma)^2
  Cp  <- (SSE / MSE) - (n0 - 2 * (n0 - m$df))
  AIC <- n0 * log(SSE) - n0 * log(n0) + 2 * (n0 - m$df)
  SBC <- n0 * log(SSE) - n0 * log(n0) + (log(n0)) * (n0 - m$df)
  X <- model.matrix(m)
  H <- X %*% solve(t(X) %*% X) %*% t(X)
  d <- e / (1 - diag(H))
  PRESS <- t(d) %*% d
  tbl <- as.data.frame(cbind(SSE = SSE, R.sq = R.sq, R.adj = R.adj,
                             Cp = Cp, AIC = AIC, SBC = SBC, PRD = PRESS))
  names(tbl) <- c("SSE", "R.sq", "R.adj", "Cp", "AIC", "SBC", "PRESS")
  tbl
}

output.sum <- rbind(select(full.model), select(reduced.model))
row.names(output.sum) <- c("full.model", "reduced.model")
kable(output.sum, digits = 3, caption = "Goodness-of-fit Measures of Candidate Models")
```

The reduced model has a slightly lower R² at 0.602, but a
nearly identical adjusted R² at 0.591 and 0.596, and clearly better AIC,
SBC, and PRESS than the full model, since it achieves almost the same
fit with far fewer parameters. After comparing the two models, **we choose the reduced model as our final model.**

## Final Model
In this section we will finalize our model after reducing variable and confirming normality assumptions

```{r final-model}
kable(summary(reduced.model)$coef, digits = 4,
      caption = "Inferential Statistics of Final Model")
```

We can explicitly write the final model as

$$
\widehat{write} = 11.63 + 0.13\times read + 0.24\times math + 0.24\times science + 0.23\times socst - 5.49\times gender_{male}
$$

Since none of the variables were transformed, every coefficient has a
direct practical interpretation. Holding the other test scores and
gender constant, a one-point increase in math score is associated with
about a 0.24-point increase in writing score; science and social
studies scores show similar associations (about 0.23-0.24 points each).
Reading score has a smaller association (0.13 points) and is only
borderline significant (p ≈ 0.055) once the other three test scores are
already in the model, since they are all correlated with each other.
Male students are predicted to score about 5.49 points lower on writing
than female students, holding all test scores constant - the single
largest effect in the model.

# Bootstrap Confidence Intervals
In this section we construct bootstrap confidence intervals for the model.

```{r bootstrap-mlr}
B <- 5000
n <- nrow(hsb2)

boot.coefs <- matrix(NA, nrow = B, ncol = length(coef(reduced.model)))
colnames(boot.coefs) <- names(coef(reduced.model))

for (i in 1:B) {
  boot.id  <- sample(1:n, n, replace = TRUE)
  boot.fit <- lm(write ~ read + math + science + socst + gender,
                  data = hsb2[boot.id, ])
  boot.coefs[i, ] <- coef(boot.fit)
}

boot.ci <- t(apply(boot.coefs, 2, quantile, probs = c(0.025, 0.975)))
pander(boot.ci, caption = "95% Bootstrap Confidence Intervals - Reduced MLR Model")
```

```{r compare-ci-mlr}
ols.ci <- confint(reduced.model, level = 0.95)
comparison <- data.frame(
  Variable         = rownames(ols.ci),
  OLS_Lower        = round(ols.ci[, 1], 4),
  OLS_Upper        = round(ols.ci[, 2], 4),
  Bootstrap_Lower  = round(boot.ci[, 1], 4),
  Bootstrap_Upper  = round(boot.ci[, 2], 4)
)
pander(comparison, caption = "95% Confidence Intervals: OLS vs. Bootstrap")
```

The OLS and bootstrap confidence intervals agree closely across every
coefficient, which is expected given that the reduced model's residuals
already satisfied the normality assumption. 

# Discussion

We started with a full model including every available predictor,
checked for multicollinearity with VIF and used
significance testing to drop four insignificant categorical predictors
(`race`, `ses`, `schtyp`, `prog`). This resulted in a simpler model with variables
`read`, `math`, `science`, `socst`, and `gender`. The reduced model's
residuals passed a formal normality test, but a visible bend remained in
the red lines in the Residuals vs Fitted and Scale-Location plots. We checked this
directly with a Box-Cox transformation. The transformation produced a
negligible improvement in fit and normality, and the bend itself
was still there. Therefore we kept the model in its original scale. The
goodness-of-fit comparison supported keeping the simpler model over the
full one, and the bootstrap confidence intervals closely matched the OLS
results, reinforcing that math score, science score, social studies
score, and gender are all reliably associated with writing score, with
gender showing the largest single effect.

Our practical question asked which factors are associated with a
student's writing score. The final model directly answers this: math
score (p = 0.0005), science score (p = 0.0001), social studies score
(p < 0.0001), and gender (p < 0.0001) are all statistically significant
predictors of writing score at the $\alpha = 0.05$ level, while reading
score (p = 0.0550) falls just short of conventional significance once
the other three test scores are already accounted for. This does not
mean reading ability is unrelated to writing ability - the pairwise
scatter plot in the exploratory analysis showed a clear positive
correlation between the two (r = 0.60) - it means that once math,
science, and social studies scores are already in the model, reading
score adds comparatively little *additional* explanatory power, since
all four test scores are correlated with one another.

Gender is the single strongest and most reliable predictor: male
students are predicted to score about 5.49 points lower on writing than
female students (95% CI: 3.78 to 7.22 points lower, from both the OLS
and bootstrap intervals), holding all test scores constant. Because both
inference methods agree so closely, we have strong evidence this
difference is real rather than a byproduct of any residual assumption
violation.

Notably, none of the categorical variables originally hypothesized to
matter for this question - socioeconomic status (`ses`) and academic
program (`prog`) - were statistically significant once test scores and
gender were accounted for (both were removed during variable selection).
This means that, at least for this data set, a student's own academic
performance and gender are more directly tied to writing score than
their socioeconomic background or program track. The model overall
explains about 60.2% of the variation in writing scores ($R^2 = 0.602$),
indicating other unmeasured factors also play a meaningful role.