---
title: "Highschool and Beyond survey analysis"
author: "Joshua Xavier"
date: "2026-09-14"
output: 
  html_document:
    code_folding: hide
  
---

```{=html}
<style type="text/css">

/* Cascading Style Sheets (CSS) is a stylesheet language used to describe the presentation of a document written in HTML or XML. it is a simple mechanism for adding style (e.g., fonts, colors, spacing) to Web documents. */

h1.title {  /* Title - font specifications of the report title */
  font-size: 24px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}
h4.author { /* Header 4 - font specifications for authors  */
  font-size: 20px;
  font-weight: bold;
  font-family: system-ui;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - font specifications for the date  */
  font-size: 18px;
  font-weight: bold;
  font-family: system-ui;
  color: DarkBlue;
  text-align: center;
  font-weight: bold;
}
h1 { /* Header 1 - font specifications for level 1 section title  */
    font-size: 22px;
    font-family: system-ui;
    color: navy;
    text-align: left;
    font-weight: bold;
}
h2 { /* Header 2 - font specifications for level 2 section title */
    font-size: 20px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
    font-weight: bold;
}

h3 { /* Header 3 - font specifications of level 3 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
    font-weight: bold;
}

h4 { /* Header 4 - font specifications of level 4 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
    font-weight: bold;
}

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

</style>
```

```{r setup, include=FALSE}
# Detect, install and load packages if needed.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("MASS")) {
   install.packages("MASS")
   library(MASS)
}
if (!require("nleqslv")) {
   install.packages("nleqslv")
   library(nleqslv)
}
#
if (!require("pander")) {
   install.packages("pander")
   library(pander)
}

if (!require("psych")) {   
  install.packages("psych")
   library(psych)
}
if (!require("MASS")) {   
  install.packages("MASS")
   library(MASS)
}

knitr::opts_chunk$set(echo = FALSE,  # include code chunk in the output file
                   warning = FALSE,  # Sometimes, your code may produce a warning
                                     # messages, you can choose to include the
                                     # warning messages in the output file. 
                   message = FALSE,  
                   results = TRUE,   # you can also decide whether to include 
                                     # the output in the output file.
                   comment = FALSE   # Suppress hash-tags in the output results.
                      )   
library(tidyverse)
library(knitr)
library(psych)
```

# Data set and Practical Questions

```{r import-data}
hsb2 <- read.csv("https://raw.githubusercontent.com/Joshuaxav/STA321F26/main/hsb2.csv")
#glimpse(hsb2)
```

## Description of data set
```{r Intro}
var_table <- tibble(
  Variable = c("id", "write", "read", "math", "science", "socst",
               "gender", "race", "ses", "schtyp", "prog"),
  Type = c("Numerical (identifying)",
           "Numerical (response candidate, continuous)",
           "Numerical (explanatory)",
           "Numerical (explanatory)",
           "Numerical (explanatory)",
           "Numerical (explanatory)",
           "Categorical (2 levels)",
           "Categorical (4 levels)",
           "Categorical (3 levels)",
           "Categorical (2 levels)",
           "Categorical (3 levels)"),
  Description = c("Student id number",
                  "Writing test score",
                   "Reading test score",
                   "Math test score",
                   "Science test score",
                   "Social studies test score",
                   "Student gender",
                   "Student race/ethnicity",
                   "Socioeconomic status: low/middle/high",
                   "School type: public/private",
                   "Program type: general/academic/vocational")
)
#points from package (knitr) to the function.
#Keeps things organized and less confusing
knitr::kable(var_table)
```
The data set used in this project is the "hsb2" data set, an acronym for "high school and beyond". The data set contains 200 randomly selected observations from the "High School and Beyond" survey, which was conducted by the National Center for Education Statistics and polled seniors in high school. The dataset uses 11 total variables, 1 identifying variable (id), 5 numerical variables, including the response variable write, and 5 categorical variables.

The dataset contains enough information to investigate these questions because it includes socioeconomic status, academic program, demographic characteristics, and multiple standardized test scores for 200 students, with 11 variables. This allows us to compare academic preformance across different groups and use statistical methods to determine whether a meaningful relationships exist.

## Practical Question 
Using this data set, we hope to find whether a students socioeconomic status and academic program relate with their preformance in highschool. Analytically, we would like to see whether there is a statistically significant relationship between socioeconomic status and a students standardized writing scores. we would also like to see whether writing scores can be predicted using socioeconomic status or other test scores. 

# Data Analysis
First we explore the pairwise association between the numerical variables
in the data set.

## Pairwise Scatter Plot
```{r pairwise-scatterplot, fig.align='center', fig.width=7, fig.height=7, fig.cap="Figure 1: Pairwise scatter plot."}
pairs.panels(hsb2[, c("read", "write", "math", "science", "socst")],
  main = "Pairwise Scatterplots of Numerical Variables",
             pch = 21,
             bg = "lightblue",
             )

```
The five variables all show a positive relationship with each other. In general, students who scored higher in one subject also scored higher in the others. Reading and math had the strongest relationship at 0.66, while science and social studies had the weakest at 0.47. It is also worth mentioning reading had a particularly strong relationship with the other subjects. 

The points follow a fairly straight upward pattern, so a linear model appears to be appropriate for predicting writing scores from reading scores.

If a transformation is needed, we would transform the response variable. This is because
the Box Cox transformation helps fix stabilize variance, normalize residuals, and improves linearity, which are tied to the response variable in a linear regression model. 

# Fitting SLR to Data 
We use parametric and bootstrap regression models to assess the
association between writing score and reading score

## Scatter Plot
```{r scatterplot, fig.align='center', fig.width=5, fig.height=5, fig.cap=" Figure 2: Scatter plot between reading and writing scores."}
write <- hsb2$write
reading <- hsb2$read
plot(reading, write,
     pch = 21,
     bg = "lightblue",
     col = "navy",
     main = "Relationship Between Reading and Writing Scores",
     xlab = "Reading Score",
     ylab = "Writing Score",

abline(lm(write ~ reading), lwd = 2)
)
```
Figure 2 shows a positive relationship between writing test scored and reading test scores.

## Parametric Simple Linear Regression
```{r parametric-fit, fig.align='center', fig.height=7, fig.width=5}
library(broom)
parametric.model <- lm(write ~ reading, data = hsb2)
par(mfrow = c(2,2))
plot(parametric.model)
tidy_model <- tidy(parametric.model)
kable(tidy_model, digits = 4,
      caption = paste0("Coefficient estimates (R² = ",
                        round(summary(parametric.model)$r.squared, 3), ")"))
       
```

### Residual Plots

We can see from the residual plots: 

* The Residuals Vs Fitted plot has points randomly scattered points, but a bent line possibly indicating a linearity violation.

* The Scale-Location plot, similar to the vs fitted plot, has points scattered randomly, but a bent line possibly indicating the constant variance assumption is not satisfied

* The QQ Residual plot shows departures from the red line near both tails, suggesting a violation of the normality assumption.

* The Residuals Vs Leverage plot indicates that there are no serious outliers,

From the SLR model, we see a positive relationship between reading and writing scores, with a reading coefficient of 0.55. This means that for every one point increase in reading score, the predicted writing score increases by about 0.55 points. The R² value is 0.356, meaning that reading scores explain about 35.6% of the variation in writing scores. The p-value for reading is less than 2.2 × 10⁻¹⁶, providing very strong evidence of a statistically significant relationship between reading and writing scores.

## Box Cox Transformation
Since the residual plots show possible violations, we check if transformations would
fix the problems to a meaningfull degree.

```{r boxcox-plot, fig.align='center', fig.width=5, fig.height=5, fig.cap="Box-Cox transformation."}
boxcox(lm(write ~ reading, data = hsb2), lambda = seq(1, 3, 1/10))
```

```{r boxcox-apply, fig.align='center', fig.width=7, fig.height=7, fig.cap="Box-Cox transformed SLR."}
lambda <- 2.3
write.bc <- (write^lambda - 1) / lambda
boxcox.m <- lm(write.bc ~ reading)
par(mfrow = c(2, 2))
plot(boxcox.m)
```
```{r boxcox-compare}
compare_table <- tibble(
  Model = c("write ~ read (raw)", "Box-Cox transformed (lambda = 2.3)"),
  R_squared = c(summary(parametric.model)$r.squared, summary(boxcox.m)$r.squared),
  Shapiro_p = c(shapiro.test(residuals(parametric.model))$p.value,
                shapiro.test(residuals(boxcox.m))$p.value)
)
knitr::kable(compare_table, digits = 4)
```
The transformed model only shows small improvement with the R² rising slightly
from 0.356 to 0.367. The p-value for residual normality also rises from 0.007 to 0.039. Despite this, The residual plots still show violations after transformation. The Residuals Vs Fitted plot has a bent line possibly indicating a linearity violation. The Scale-Location plot has a bent line indicating constant variance is not satisfied. Along with this, the QQ Residual plot shows departures from line near both tails, suggesting a violation of the normality assumption. Therefore, we choose not to transform and stick with the model in its original scale, `write ~ read`. 


## Bootstrap confidence intervals
```{r Bootstrap confidence intervals}

B <- 5000
vec.id <- 1:length(write)

boot.beta0 <- NULL
boot.beta1 <- NULL

for(i in 1:B){
  boot.id <- sample(vec.id, length(write), replace = TRUE)
  
  boot.reg <- lm(write[boot.id] ~ reading[boot.id])
  
  boot.beta0[i] <- coef(boot.reg)[1]
  boot.beta1[i] <- coef(boot.reg)[2]
}

boot.beta0.ci <- quantile(boot.beta0, c(0.025, 0.975), type = 2)
boot.beta1.ci <- quantile(boot.beta1, c(0.025, 0.975), type = 2)

boot.coef <- data.frame(
  Intercept = boot.beta0.ci,
  Reading = boot.beta1.ci
)

rownames(boot.coef) <- c("2.5%", "97.5%")

pander(boot.coef,
       caption = "95% Bootstrap Confidence Intervals for Regression Coefficients")

```


The bootstrap confidence interval for the reading coefficient is (0.464 to 0.644). Because this interval does not contain zero, it provides strong evidence that reading score is related to writing score. This agrees with the ordinary least squares regression, which produced a p-value = 2.2 × 10⁻¹⁶. The bootstrap confidence interval also provides a useful check because the residual diagnostics showed some possible violations of the regression assumptions. I would report the bootstrap confidence interval along with the raw regression results because it provides an additional, less assumption-dependent measure of uncertainty for the regression coefficient.


```{}
