---
title: 'STA321: Week #02 Assignment'
author: "Joshua Xavier"
date: "2026-09-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Question 2

```{r Question 2}
library(tidyverse)
#setwd("C:/Users/joshu/OneDrive/Desktop/Fall 26/STA 321")
protein <- read_csv("w02-Protein_Supply_Quantity_Data.csv")
#View(protein)
Spices <- protein$Spices
n <- length(Spices)
xbar <- mean(Spices)
med <- median(Spices)
```

For this assignment I am using the **Spices** variable from the protein
supply data set. This variable measures the percentage of protein for each of 
the `r n` countries with a recorded value.
Since spices are typically a minor part of meal and diet, I expect this variable to show right skew with most countries being at almost 0.

## Question 3
```{r Question 3}
s <- sd(Spices)
se <- s / sqrt(n)
t.crit <- qt(0.975, df = n - 1)

lower.t <- xbar - t.crit * se
upper.t <- xbar + t.crit * se
c(lower.t, upper.t)


```
The traditional 95% confidence interval for the mean of Spices is
(`r round(lower.t, 4)`, `r round(upper.t, 4)`).


## Question 4
```{r Question 4}


original.sample = sample( Spices, 170, replace = FALSE)
spiceSample = NULL
for(i in 1:1000){   
  
 replace = sample(x = original.sample, size = 170, replace = TRUE)
 spiceSample[i] = mean(replace)
}

spiceSample.CI <- quantile(spiceSample, c(0.025, 0.975))
spiceSample.CI

```
The bootstrap 95% confidence interval for the mean of Spices is
(`r round(spiceSample.CI[1], 4)`, `r round(spiceSample.CI[2], 4)`).



## Question 5
```{r Question 5}
hist(spiceSample)
```

## Question 6
```{r Question 6}

```
The traditional interval is (`r round(lower.t, 4)`, `r round(upper.t, 4)`) and 
the bootstrap interval is (`r round(spiceSample.CI[1], 4)`, `r round(spiceSample.CI[2], 4)`). 
The two intervals are almost the same and are centered around the 
same value, but the bootstrap interval is not perfectly symmetric around the 
mean like the traditional interval is. This makes sense given the right skew 
in the Spices variable. The traditional method assumes the sampling 
distribution of the mean is normal, while the bootstrap method 
does not, which shows in the skewness of the 
data through resampling.

We are 95% confident that the true mean percentage of protein supply coming 
from spices, across countries, lies between `r round(spiceSample.CI[1], 4)` 
and `r round(spiceSample.CI[2], 4)`, based on the bootstrap method.
